Parallel algorithm for solving multicriterial optimization problems using elements of machine learning

### 2025-06-17 2025/RSCD_MCO/RSD-2025.tex
fix rev. comments

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}

\usepackage{amsmath}
\usepackage{marvosym}
\usepackage{hyperref}

@@

\begin{figure}
\center
\includegraphics[width=0.7\textwidth]{fig1.png}
\caption{Computational scheme of the algorithm for solving multicriterial optimization problems using machine learning models.} \label{fig1}
\end{figure}

@@

In the framework of the developed approach \cite{Konnov2025}, the second option with scalarization of criteria is used. To build a good estimate of the Pareto set, a family of problems
\begin{equation}
\label{eq:04}
\min {\varphi_j (y)}=F(\lambda_j,y),F(\lambda_j,y) \in \{F(\lambda_1,y),F(\lambda_2,y),\dots,F(\lambda_\tau,y)\},y\in D,
\end{equation}
is solved,  where $F(\lambda_j,y)$, $1 \leq j \leq \tau$, is a multiextremal function, obtained on the base of scalarization of criteria $f_i(y)$, $1 \leq i \leq s$. Here $\lambda_j=\{\lambda_{i,j}\}$, $1 \leq i \leq s$, $1 \leq j \leq \tau$, are significance indicators for each criterion, s is the number of criteria, $\tau$ is the number of scalar optimization problems. There are different ways to build functions $F(\lambda_j,y)$ \cite{Miettinen1999,Konnov2025,Marler2004}. The performed computational experiments used minimax convolution of criteria
\begin{equation}
\label{eq:05}
F(\lambda_j, y) = \max_{1 \leq i \leq s} \left(\lambda_{i,j} f_i (y)\right), y \in D, \sum_{i=1}^s {\lambda_{i,j}} = 1, 0 \leq \lambda_{i,j} \leq 1.
\end{equation}

@@

\begin{figure}
\center
\includegraphics[width=0.6\textwidth]{fig2.png}
\caption{Example of Pareto set evaluation in solving two scalar optimization problems (trial points are presented in the criteria space).} \label{fig2}
\end{figure}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Neural network models used to calculate the characteristic $R_{PS}(i)$.} \label{fig3}
\end{figure}

@@

\begin{figure}
\center
\includegraphics[width=0.6\textwidth]{fig4.png}
\caption{Examples of Pareto set evaluation for the case of two scalar optimization problems.} \label{fig4}
\end{figure}

@@

Computational experiments were carried out on the nodes of the Lobachevsky supercomputer of Nizhny Novgorod State University with two processors Intel Xeon Silver 4310T 2.3 GHz, 64 GB RAM. Machine learning libraries SciKit-Learn 0.24.2 \cite{scikit-learn} and PyTorch 2.5.1 \cite{PyTorch} were used in experiments. The algorithm proposed in the work was implemented in Python using the framework iOpt \cite{ioptmco}.

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of HV index values for sequential algorithm implementation.} \label{fig5}
\end{figure}

@@

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}
\begin{table}[ht]
\centering
\caption{Average number of trials per process}
\label{tab:1}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
\textbf{Number of processes}            & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}    & \textbf{20}    \\ \hline
MGSA                                    & 2269.82         & 1385.02         & 746.57          & 373.36          & 202.01            & 143.52         \\
\textbf{MGSA   Linear SVC,  $\alpha$ = 0.05}   & \textbf{686.13} & \textbf{408.79} & \textbf{213.6}  & \textbf{111.85} & \textbf{70.95} & \textbf{62.16} \\
MGSA Poly SVC, $\alpha$ = 0.05               & 909.59          & 544.19          & 278.62          & 153.64          & 95.37          & 77.94          \\
MGSA RBF SVC, $\alpha$ = 0.04                & 652.74          & 446.84          & 233.81          & 119.84          & 72.36          & 62.37          \\
\textbf{MGSA Dist, $\alpha$ = 0.01}          & \textbf{659.98} & \textbf{402.9}  & \textbf{211.33} & \textbf{112.38} & \textbf{70.79} & \textbf{62.00} \\
MGSA NN 1 layer, $\alpha$ = 0.05             & 795.16          & 446.25          & 232.08          & 117.16          & 75.19          & 62.70          \\
\textbf{MGSA NN 2 layers, $\alpha$ =   0.03} & \textbf{682.12} & \textbf{401.74} & \textbf{219.62} & \textbf{113.32} & \textbf{73.51} & \textbf{61.32} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Average HV Index in dependence on number of processes}
\label{tab:2}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
Number of processes                   & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}     & \textbf{20}     \\ \hline
\textbf{MGSA}                         & \textbf{114.91} & \textbf{114.79} & \textbf{114.84} & \textbf{114.90} & \textbf{115.00} & \textbf{114.96} \\
MGSA Linear SVC, $\alpha$ = 0.05           & 114.44          & 114.65          & 114.72          & 114.74          & 114.91          & 114.96          \\
\textbf{MGSA   Poly SVC, $\alpha$ = 0.05} & \textbf{114.61} & \textbf{114.64} & \textbf{114.71} & \textbf{114.8}  & \textbf{114.95} & \textbf{115.01} \\
MGSA RBF SVC, $\alpha$ = 0.04           & 114.54          & 114.54          & 114.73          & 114.89          & 114.92          & 114.96          \\
MGSA Dist, $\alpha$ = 0.01                 & 114.66          & 114.55          & 114.69          & 114.73          & 114.92          & 114.93          \\
MGSA NN 1 layer, $\alpha$ = 0.05           & 114.77          & 114.61          & 114.72          & 114.81          & 114.96          & 114.98          \\
\textbf{MGSA NN 2 layers, $\alpha$ = 0.03} & \textbf{114.74} & \textbf{114.54} & \textbf{114.71} & \textbf{114.76} & \textbf{114.98} & \textbf{115.01} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig6.png}
\caption{Dependence of the mean value of the index HV from the trial number per process for the case of 20 processes.} \label{fig6}
\end{figure}

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig7.png}
\caption{Dependence of the HV index on the number of trials executed for different number of processes.} \label{fig7}
\end{figure}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{marvosym}
\usepackage{hyperref}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.6\textwidth]{fig1.png}
\caption{Computational scheme of the algorithm for solving multicriterial optimization problems using machine learning models.} \label{fig1}
\end{figure}

@@

In the framework of the developed approach \cite{Konnov2025}, the second option with scalarization of criteria is used. To build a good estimate of the Pareto set, a family of problems
\begin{equation}
\label{eq:04}
\min {\varphi_j (y)}=F(\lambda_j,y),F(\lambda_j,y) \in \{F(\lambda_1,y),F(\lambda_2,y),\dots,F(\lambda_\tau,y)\},y\in D,
\end{equation}
is solved,  where $F(\lambda_j,y)$, $1 \leq j \leq \tau$, is a multiextremal function, obtained on the base of scalarization of criteria $f_i(y)$, $1 \leq i \leq s$. Here $\lambda_j=\{\lambda_{i,j}\}$, $1 \leq i \leq s$, $1 \leq j \leq \tau$, are significance indicators for each criterion, s is the number of criteria, $\tau$ is the number of scalar optimization problems. There are different ways to build functions $F(\lambda_j,y)$ \cite{Miettinen1999,Konnov2025,Marler2004}.  
%В рамках предлагаемого алгоритма $\tau$ фиксируеться. Множество наборов $\lambda_j$, $1 \leq j \leq \tau$, выбирается случайно с использованием генератора Соболя в множестве  возможных значений (\ref{eq:051}). 
{\color{red} In the proposed algorithm, $\tau$ is fixed. The set of $\lambda_j$, $1 \leq j \leq \tau$, is chosen randomly using the Sobol generator\cite{Sobol2003} in the space of possible values (\ref{eq:051})}

\begin{equation}
\label{eq:051}
\sum_{i=1}^s {\lambda_{i,j}} = 1, 0 \leq \lambda_{i,j} \leq 1.
\end{equation}

The performed computational experiments used minimax convolution of criteria

\begin{equation}
\label{eq:05}
F(\lambda_j, y) = \max_{1 \leq i \leq s} \left(\lambda_{i,j} f_i (y)\right), y \in D.
\end{equation}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.7\textwidth]{fig2.png}
\caption{Example of Pareto set evaluation in solving two scalar optimization problems (trial points are presented in the criteria space).} \label{fig2}
\end{figure}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Neural network models used to calculate the characteristic $R_{PS}(i)$.} \label{fig3}
\end{figure}

@@

\begin{figure}[t]
\center
\includegraphics[width=\textwidth]{fig4.png}
\caption{Examples of Pareto set evaluation for the case of two scalar optimization problems.} \label{fig4}
\end{figure}

@@

Computational experiments were carried out on the nodes of the Lobachevsky supercomputer of Nizhny Novgorod State University with two processors Intel Xeon Silver 4310T 2.3 GHz, 64 GB RAM. Machine learning libraries SciKit-Learn 0.24.2 \cite{scikit-learn} and PyTorch 2.5.1 \cite{PyTorch} were used in experiments. The algorithm proposed in the work was implemented in Python using the framework iOpt \cite{ioptmco}. 
%Сравнение предыдущей последовательной версии предлагаемого алгоритма с современными разработками можно найти в нашей статье [Konnov2025]. В этом разделе больше внимания уделено анализу эффективности параллельной реализации.
{\color{red}  A comparison of the previous sequential version of the proposed algorithm with state-of-the-art developments can be found in our paper \cite{Konnov2025}. This section focuses more on the analysis of the efficiently of parallel implementation.}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of HV index values for sequential algorithm implementation.} \label{fig5}
\end{figure}

@@

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}
\begin{table}[t]
\centering
\caption{Average number of trials per process}
\label{tab:1}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
\textbf{Number of processes}            & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}    & \textbf{20}    \\ \hline
MGSA                                    & 2269.82         & 1385.02         & 746.57          & 373.36          & 202.01            & 143.52         \\
\textbf{MGSA   Linear SVC,  $\alpha$ = 0.05}   & \textbf{686.13} & \textbf{408.79} & \textbf{213.6}  & \textbf{111.85} & \textbf{70.95} & \textbf{62.16} \\
MGSA Poly SVC, $\alpha$ = 0.05               & 909.59          & 544.19          & 278.62          & 153.64          & 95.37          & 77.94          \\
MGSA RBF SVC, $\alpha$ = 0.04                & 652.74          & 446.84          & 233.81          & 119.84          & 72.36          & 62.37          \\
\textbf{MGSA Dist, $\alpha$ = 0.01}          & \textbf{659.98} & \textbf{402.9}  & \textbf{211.33} & \textbf{112.38} & \textbf{70.79} & \textbf{62.00} \\
MGSA NN 1 layer, $\alpha$ = 0.05             & 795.16          & 446.25          & 232.08          & 117.16          & 75.19          & 62.70          \\
\textbf{MGSA NN 2 layers, $\alpha$ =   0.03} & \textbf{682.12} & \textbf{401.74} & \textbf{219.62} & \textbf{113.32} & \textbf{73.51} & \textbf{61.32} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{table}[t]
\centering
\caption{Average HV Index in dependence on number of processes}
\label{tab:2}
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{ccccccc}
\hline
Number of processes                   & \textbf{1}      & \textbf{2}      & \textbf{4}      & \textbf{8}      & \textbf{16}     & \textbf{20}     \\ \hline
\textbf{MGSA}                         & \textbf{114.91} & \textbf{114.79} & \textbf{114.84} & \textbf{114.90} & \textbf{115.00} & \textbf{114.96} \\
MGSA Linear SVC, $\alpha$ = 0.05           & 114.44          & 114.65          & 114.72          & 114.74          & 114.91          & 114.96          \\
\textbf{MGSA   Poly SVC, $\alpha$ = 0.05} & \textbf{114.61} & \textbf{114.64} & \textbf{114.71} & \textbf{114.8}  & \textbf{114.95} & \textbf{115.01} \\
MGSA RBF SVC, $\alpha$ = 0.04           & 114.54          & 114.54          & 114.73          & 114.89          & 114.92          & 114.96          \\
MGSA Dist, $\alpha$ = 0.01                 & 114.66          & 114.55          & 114.69          & 114.73          & 114.92          & 114.93          \\
MGSA NN 1 layer, $\alpha$ = 0.05           & 114.77          & 114.61          & 114.72          & 114.81          & 114.96          & 114.98          \\
\textbf{MGSA NN 2 layers, $\alpha$ = 0.03} & \textbf{114.74} & \textbf{114.54} & \textbf{114.71} & \textbf{114.76} & \textbf{114.98} & \textbf{115.01} \\ \hline
\end{tabular}%
}
\end{table}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.8\textwidth]{fig6.png}
\caption{Dependence of the mean value of the index HV from the trial number per process for the case of 20 processes.} \label{fig6}
\end{figure}

@@

\begin{figure}[t]
\center
\includegraphics[width=0.8\textwidth]{fig7.png}
\caption{Dependence of the HV index on the number of trials executed for different number of processes.} \label{fig7}
\end{figure}
